The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-11 paper/iphas-dr2.tex
-

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering the Galactic latitude range $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against the Sloan Digital Sky Survey (SDSS),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a uniform calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration}
\footnote{The term ubercalibration refers to the name of the code
first used to re-calibrate SDSS photometry using repeat observations,
and is a term that has since been adopted by other surveys.
The name is an anglicised version of the German word \"uberkalibration,
which was reportedly chosen because authors Doug Finkbeiner and David Schlegel
have German-sounding names (Finkbeiner 2010).} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep[e.g.][]{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{./plots/errors/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue, 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These uncertainties are based only on the (Poissonian)
             photon noise and hence this figure does not show
             systematic or calibration uncertainties.}
    \label{fig:uncertainties}
\end{figure}

@@

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/pairmag.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/pairmag-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for both all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         The values shown are the mean absolute residuals
         between the primary and the secondary detections,
         while the error-bars show the standard deviations.
         Left panel: 
         the best photometric repeatability is reached at $r=14$
         with $\sigma_{r-r2}~=~0.041$~mag.
         Bright stars at $r<13$ and $i<12$ 
         show increasing uncertainties due to saturation effects.
         Right panel: we find that applying the quality criteria
         has improved the photometric repeatability significantly.
         The best repeatability is again reached at $r=14$
         but has reduced to $\sigma=0.028$ mag.
         The quality criteria have also been successful
         at removing objects with large systematics at the bright
         and faint ends.
    }
\end{figure*}

@@

The target of our demonstration is Sh 2-82:
a 5~arcmin-wide H{\sc ii} region located near $(l,b)=(53.55^\circ, 0.00^\circ)$
in the constellation of Sagitta.
Nicknamed by amateur astronomers as the ``Little Cocoon Nebula'',
Sh 2-82 is ionised by 
the $\sim$10th magnitude star HD\,231616
% Georgelin1973 claims B0III, others claim B0V
with spectral type B0V/III
\citep{Georgelin1973,Mayer1973,Hunter1990}.
This ionising star has been placed 
at a likely distance of 1.5-1.7 kpc
based on its photometric parallax
\citep{Mayer1973,Lahulla1985,Hunter1990}.

@@

\appendix
\onecolumn
\section{Catalogue format}
\label{app:columns}

>>>>

\restylefloat{table}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering the Galactic latitude range $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to mean $5\sigma$-depths of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a uniform calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration}\footnote{`Ubercalibration'
refers to the name of the code used to re-calibrate SDSS photometry
using repeat observations. 
This name -- which has since been adopted by other surveys --
is an anglicised version of the German word `\"uberkalibration',
which was reportedly chosen because authors Doug Finkbeiner and David Schlegel
both have German-sounding names (Finkbeiner 2010).} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep[e.g.][]{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/uncertainties/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue, 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These uncertainties are based only on the (Poissonian)
             photon noise and hence this figure does not show
             the level of systematic uncertainties.}
    \label{fig:uncertainties}
\end{figure}

@@

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/repeatability.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/repeatability-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for both all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         The values shown are the mean absolute residuals
         between the primary and the secondary detections,
         while the error-bars show the standard deviations.
         Left panel: 
         the best photometric repeatability is reached at $r=14$
         with $\sigma_{r-r2}~=~0.041$~mag.
         Bright stars at $r<13$ and $i<12$ 
         show increasing uncertainties due to saturation effects.
         Right panel: we find that applying the quality criteria
         has improved the photometric repeatability significantly.
         The best repeatability is again reached at $r=14$
         but has reduced to $\sigma=0.028$ mag.
         The quality criteria have also been successful
         at removing objects with large systematics at the bright
         and faint ends.
    }
\end{figure*}

@@

The target of our demonstration is Sh 2-82:
a 5~arcmin-wide H{\sc ii} region located near $(l,b)=(53.55^\circ, 0.00^\circ)$
in the constellation of Sagitta.
Nicknamed by amateur astronomers as the ``Little Cocoon Nebula'',
Sh 2-82 is ionised by 
the $\sim$10th magnitude star HD\,231616
with spectral type B0V/III
\citep{Georgelin1973,Mayer1973,Hunter1990}.
This ionising star has been placed 
at a likely distance of 1.5-1.7 kpc
based on its photometric parallax
\citep{Mayer1973,Lahulla1985,Hunter1990}.

@@

\newpage
\appendix

\section{Converting IPHAS magnitudes into fluxes}

Mean photon wavelength of the passband following eqn.~A14 in Bessel \& Murphy (2012):
\begin{equation}
\lambda_0 = \frac{\int{\lambda S(\lambda) d \lambda}}
{\int{S(\lambda)d\lambda}}
\end{equation}

Pivot wavelength of the passband following eqn.~A16 in Bessel \& Murphy (2012):
\begin{equation}
\lambda_p = \frac{\int{S(\lambda) \lambda d \lambda}}
{\int{\frac{S(\lambda)}{\lambda} d\lambda}}
\end{equation}

Mean monochromatic flux,
following eqn.~A11 in Bessel \& Murphy (2012):
\begin{equation}
f_\lambda = \frac{\int{f_\lambda(\lambda) S(\lambda) \lambda d \lambda}}
{\int{S(\lambda)\lambda d \lambda}}
\end{equation}
where $S(\lambda)$ includes the filter, atmosphere and QE.

\begin{table}[h]
    \caption{Wavelength and mean monochromatic flux density for Vega in the IPHAS filter system.}
    \label{tbl:flux}
    \begin{tabular}{lrrrr}
    \toprule
     Band & $\lambda_0$ & $\lambda_p$ & EW  & $f_\lambda$   \\
     & \AA & \AA & \AA & erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$  \\
    \midrule
$r$ &  6223  &   6211  &  785.6  &  2.47e-9  \\
\ha &  6568  &   6568  &   59.6  &  1.81e-9  \\
$i$ &  7674  &   7674  &  759.9  &  1.30e-9   \\
    \bottomrule
    \end{tabular}
\end{table}

\newpage
\onecolumn
\section{Catalogue format}
\label{app:columns}
